---
title: "Palau's Coral Carbonate Production and Relation to Growth Potential"
author: "Emily Buono"
date: "7/17/2020"
output: html_document
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Coral's Carbonate Production Rates around Palau 


## Research Question
What is the Net Carbonate Production of multiple species of Coral at several research sites around Palau, and how does this relate back to the coral's growth potential?

Coral reefs play important parts in ocean environment, as well as acting as buffers for islands from tropical storm waves. Coral is a main contributor in the production of CaCO3, which plays a key role in reef framework construction. High percentage coral cover can mean high rates of CaCO3 accumulation and possibly rapid reef growth potential (Cacciapaglia & van Woesik, 2018). It is unknown how coral reefs and their carbonate production will respond to sea-level rise climate change. The goal of the paper this project is based on was to measure potential net carbonate production rates across different reef habitats in Palau and Yap and figure out which habitats are more likely to keep up with sea-level rise (Cacciapaglia & van Woesik, 2018). 

Palau is a country made of several outer islands and a main island. It is located in the western Pacific Ocean. Palau will be focused on in this project due to the fact that Yap's map was not available to work with and visualize.


### Required Packages

These are the required R packages to ensure the smooth running of this project's code.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(raster)
library(dplyr)
library(tidymodels)
library(patchwork)
```


### Loading The Data 

These data utilized is from the Biological & Chemical Oceanography Data Management Office. Three different data sets are downloaded here, and a map of the Palau Islands as well. Later some of these data tables are merged to be able to use in visualizations. 

```{r, message = FALSE, cache=TRUE}

coral = read.csv('http://dmoserv3.bco-dmo.org/jg/serv/BCO-DMO/Coral_Reef_Adjustment/extension_rates.flat0?
', sep = "", na.strings = c('nd', 'encrusting'))

sites_data = read.csv('http://dmoserv3.bco-dmo.org/jg/serv/BCO-DMO/Coral_Reef_Adjustment/site_list.flat0?', sep = "", na.strings = c('nd'))

model_data = read.csv('http://dmoserv3.bco-dmo.org/jg/serv/BCO-DMO/Coral_Reef_Adjustment/model_output.flat0?',sep = "", na.strings = c('nd'))

palau = getData('GADM', country='Palau', level = 0)

head(coral)
head(sites_data)
head(model_data)
```

The data includes `r length(unique(sites_data$study_site))` study sites around Palau. 

### Filtering Data 

To plot the data of NP onto a map of Palau, the data must be filtered to just show Palau sites. Dplyr pipes were utilized in this code.

```{r, message=FALSE}
palau_sites_data <- sites_data %>% filter(country == 'Palau')
palau_model_data <- model_data %>% filter(Country == 'Palau')
```

## Figure 1: Palau Maps with Site Data

Figure 1 features a full map of Palau and several of its smaller islands. The site locations are marked as red points. Figure 2 has more centered Latitude and Longitude axes to focus in on where the studies actually occurred. The sites are marked on figure 2's green Palau map by points colored by their corresponding NP values with a viridis hue scale.

```{r,fig.width = 9, message=FALSE}
zoomed_out <- ggplot() + geom_polygon(data=palau, aes(x=long, y=lat, group=group), fill= "white" , color= "black") + 
  geom_point(data=palau_model_data, aes(x=lon, y=lat, color = 'red'), size = 0.5) + coord_quickmap() + geom_rect(data=palau_model_data, aes( xmax = 135, xmin = 134, ymin =6.5, ymax = 8.5,  alpha = 0.5), fill = NA ,color = 'blue' ) + labs( y = 'Latitude' , x = 'Longitude',  title = 'Palau Island Study Sites') + theme(axis.text.x  = element_text(angle=90), legend.position = 'none')

sites_location_map_p <- ggplot() + 
  geom_polygon(data=palau, aes(x=long, y=lat, group=group), fill= "green" , color= "black") + 
  xlim(134, 135) + ylim(6.5, 8.3)+ theme(axis.text.x  = element_text(angle=90)) + 
  geom_point(data=palau_model_data, aes(x=lon, y=lat,color = NP), size = 0.5) + 
  coord_quickmap() + theme(axis.text.x  = element_text(angle=90)) + labs( x= 'Longitude', y = 'Latitude', title = 'Palau Island Study Locations', col = 'Net Carbonate 
Production (kg CaCO3 yr-1)') + scale_colour_viridis_c()

combined_maps <- zoomed_out + sites_location_map_p + plot_layout(ncol =2) +
  plot_annotation(tag_levels = "A") 
print(combined_maps)

```

## Figure 2: Coral Species and their Average Annual Extension Rates

The Annual Extensions rates (mm/yr) were measured for each species of coral, and in this code the averages were taken and visualized. The extension rates varied greatly, showing that coral growth potential could also be affected by species as well.

```{r, message=FALSE}
coral_ext_rates = coral %>% group_by(species) %>% summarise(avg = mean(Annual_Extension)) %>% na.omit()
head(coral_ext_rates)

ggplot()+ geom_col(data= coral_ext_rates, aes(x = species, y = avg, fill = 'blue')) + labs(x = 'Species of Coral', y = ' Annual Extension rates (mm/yr)', title = '                    Average Annual Extension Rates of Coral Species') +theme_test()+ theme(axis.text.x  = element_text(angle=90),legend.position = 'none') + scale_fill_viridis_d()
```

### New Data Frame

A new data frame was made combining model_data and sites_data by using the merge function on the latitude and longitude columns. The data frame was also filtered specifically for Palau.

```{r, message=FALSE}
sites_and_models <- merge(model_data, sites_data ,by=c("lat","lon"))
palau_sites_and_models <- sites_and_models %>% filter(country == 'Palau', Country == 'Palau')
head(palau_sites_and_models)
```

## Figure 3: Net Productions at Each Site, and Colored by Reef Type

As seen in this figure, the Net Carbonate production varied greatly for each reef type, but the inner reefs seemed to stay fairly low in comparison to the outer and patch reefs, at around 1-7 kg CaCO3 yr-1.

```{r, fig.width = 6, message=FALSE}
npvssite_scatterplot <- ggplot() + geom_point(data = palau_sites_and_models, aes( x= site, y = NP, color = locat )) +
  labs( x = 'Site', y = 'Net Carbonate Production (kg CaCO3 yr-1)', title = 'Net Carbonate Production at 
Each Palau Site', col= 'Reef Types') + scale_colour_viridis_d() +theme_test()
print(npvssite_scatterplot)
```

## Figure 4: Palau Study Locations as well as NPs at Each Site

Figure 4 combines the map plot of Palau with the  scatter plot of site and Net Production of Carbonate (kg CaCO3 yr-1), giving the opportunity to compare the two visualizations, seeing where exactly the studies occurred and the corresponding carbonate production.


```{r, fig.width = 13, message=FALSE}
map_and_np_locats <- sites_location_map_p + npvssite_scatterplot + plot_layout(ncol =2) +
  plot_annotation(tag_levels = "A") 
print(map_and_np_locats)

```


## Conclusions

As seen in figure 4, the inner reefs produce on average less Net Carbonate than the outer and patch reefs. This could mean that, since coral is a main contributor in the production of CaCO3, which plays a key role in reef framework construction, those inner reefs would need a higher coral cover to make the same quantity of carbonate as the other reefs and be as sturdy as them (Cacciapaglia & van Woesik, 2018). Figure 2 shows that extension rates and growth potential can also be in part due to species. So overall the study suggests that the innermost reefs are the most vulnerable to sea-level rise and climate-change-induced stress, and the species that grow the least may be at the most risk. Inner coral species need to grow more to be able to handle the stress more effectively, but are not producing enough carbonate to be able to grow like the outer and patch reefs. The outer and patch reefs are more capable to handle the changes and grow, but still are at risk due to how rapid changes are occurring in this day and age. 

### Citations

'dplyr'  [@van2018keeping]

https://www.bco-dmo.org/dataset/737526

